---
title: "Does public transportation decrease loneliness and social isolation?"
author: "Naomi Buell and Richie Rivera"
format: html
bibliography: references.bib
freeze: true
include: false
---

## Intro

```{r}
#| label: load packages
#| echo: false

library(tidyverse)
library(janitor)
library(httr)
library(jsonlite)
library(skimr)
library(RSocrata)
library(CDCPLACES)
library(readr)
```

Load [PLACES: Local Data for Better Health, Census Tract Data 2024 release](https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-Census-Tract-D/cwsq-ngmh/about_data) using [@CDCPLACES] [@Greenlund2022]. This is 2022 data.

```{r}
#| label: get-cdc-data-with-CDCPLACES-package
dict <- get_dictionary()

cdc_places <- CDCPLACES::get_places(
    geography = "census",
    measure = c("ISOLATION", "LACKTRPT", "EMOTIONSPT"),
    age_adjust = TRUE
) |>
    clean_names() |>
    filter(measure == "Feeling socially isolated among adults")
```

Browse the CDC data.

```{r}
#| label: summ-CDC-data
# cdc_places |> summary()
# cdc_places |> skim()
cdc_places |> head()
```

```{r}
#| label: get-data-with-R-Socrata
#| include: false

## Install the required package with:
## install.packages("RSocrata")

# df <- read.socrata(
#    "https://data.cdc.gov/resource/cwsq-ngmh.json",
#    app_token = "Oor3v3DJAEoCdk17eeRa8uhFK",
#    email = "naomi.buell13@spsmail.cuny.edu",
#    password = "H@FT5%yuKwX7fw!R" # Replace with actual password
# )
```

Get [Public Transit Stops by Census Tract and ZIP Code Tabulation Area, United States, 2024](https://www.icpsr.umich.edu/web/ICPSR/studies/38605/datadocumentation#) from the National Neighborhood Data Archive (NaNDA) [@pan2023].

```{r}
#| label: get-transit

transit <- read_table("data/ICPSR_38605/DS0003/38605-0003-Data.tsv") |>
    clean_names() |>
    mutate_all(~replace(., is.na(.), 0)) # Filling the empty with 0 since they correspond with national parks
```

Browse data.

```{r}
#| label: browse-transit-data
#|
transit |> head()
# transit |> skim()
# transit |> summary()

```

Merge data.

```{r}
 data <- cdc_places |>
              rename(tract_fips20 = locationname) |>
  full_join(transit)
```

Browse merged data.

```{r}
#| label: browse-merged-data

#data |> head()
#data |> skim()
#data |> summary()

nulls_df <- data |>
  select(stateabbr, count_ntm_stops) |> 
  group_by(stateabbr) |>
  summarise(
    Null_Count = sum(is.na(count_ntm_stops)),
    Non_Null_Count = sum(!is.na(count_ntm_stops))) |> 
    arrange(desc(Null_Count))

print(nulls_df, n=100)
```

There are missing states, what are they?
```{r}
# Complete list of state abbreviations
all_states <- state.abb

# Unique, non-missing state abbreviations in your data
present_states <- unique(nulls_df$stateabbr)

# Determine missing states
missing_states <- setdiff(all_states, present_states)

# Display missing states
print(missing_states)

```

Filtering missing information. These are documented by Naomi in our Paper.
We have missing states. CT is omitted because there is no data on it from the transit stop dataset.

```{r}
df <- data |>
  mutate(density_per_sqmi = tot_pop_2020/census_tract_area20) |> # creating a population land density measure
  filter(stateabbr != "CT") |> # Removing CT
  filter(!is.na(stateabbr)) |> # Removing states where there is no data
  select(tract_fips20, count_ntm_stops, stops_per_capita, density_per_sqmi, data_value) # Selecting our columns of interest

head(df)

# 11001980000 - NATIONAL MALL   - has a pop of 17 
# 36061014300 - CENTRAL PARK    - has a pop of 1  
```

Viewing the distribution of our modeling variables

```{r}
library(ggplot2)

hist_df <- df |>
  select(count_ntm_stops, stops_per_capita, density_per_sqmi, data_value) |>
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(hist_df, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  labs(title = "Histograms of Stops Per Capita, Density Per Sqmi, and Data Value",
       x = "Value",
       y = "Count")

df |>
  arrange(desc(stops_per_capita)) |>
  head(50)
```

The field `stops_per_capita` has way too many really high values. Like there's a 12k in there when the mean value is 2.
We'll remove outliers defined as `Q3 + 1.5 * IQR`:

```{r}
get_upper_outlier_threshold <- function(data) {
  Q3 <- quantile(data, 0.75) # Third quartile
  IQR_value <- IQR(data) # Interquartile range
  upper_bound <- Q3 + 1.5 * IQR_value

  return(upper_bound)
}

hist_df <- df |>
  filter(stops_per_capita < get_upper_outlier_threshold(df$stops_per_capita)) |>
  select(count_ntm_stops, stops_per_capita, density_per_sqmi, data_value) |>
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(hist_df, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  labs(title = "Histograms of Stops Per Capita, Density Per Sqmi, and Data Value",
       x = "Value",
       y = "Count")

df |>
  arrange(desc(stops_per_capita)) |>
  head(50)
```

## Investigate: Does access to public transportation (measured by number of available stops per census tract) have impact on feeling socially isolated?

## Hypothesis test

## Data Dictionary

Feelings of social isolation among adults

:   Variable summary: Loneliness is defined as the discrepancy between a person’s desired and actual social relationships and is sometimes considered synonymous with social isolation, although they are two distinct concepts. Loneliness is an emotional response to social isolation, while social isolation is an objective measure of the lack of social interactions and relationships. Previous studies show that social isolation and loneliness are common sources of chronic stress in adults, and are common at the end of life, affecting 1 in 4 older adults. Positive relationships and interactions with family, friends, co-workers, and community members can have a protective impact on individual health and well-being, and these relationships can also help mitigate the negative impacts of challenges that people face (e.g., living in an unsafe neighborhood, trouble affording housing or food). Past research shows that people with high levels of social isolation are at higher risk of developing chronic conditions (e.g., depression, cardiovascular disease, hypertension, cancer) and have a higher risk of mortality.

Lack of reliable transportation in the past 12 months among adults

:   Nearly 2% of the U.S. population delayed medical care because they did not have transportation in 2017. Lack of available, convenient, or reliable transportation can affect a person’s ability to consistently access health care services, which can lead to delays in healthcare and medication use that can subsequently impact overall health. In addition, the inability to access reliable transportation to work, schools, and grocery stores is associated with higher rates of unemployment, poverty, and chronic illness. People with lower incomes and uninsured people are more likely to experience transportation barriers. Limited public transportation infrastructure in a neighborhood leaves people without access to vehicles lacking in reliable transportation options. Offering free or reimbursed public transit or taxi costs, connecting individuals to transportation, and policies that improve the safety and accessibility of sidewalks and bike lanes can help improve transportation access.

Lack of social and emotional support among adults

:   Positive relationships and interactions with family, friends, co-workers, and community members can have a protective impact on individual health and well-being, and these relationships can also help mitigate the negative impacts of challenges that people face (e.g., living in an unsafe neighborhood, trouble affording housing or food). Past research shows that people with high levels of social isolation (lack of interactions with others or the wider community) are at higher risk of developing chronic conditions (e.g., depression, cardiovascular disease, hypertension) and have a higher risk of mortality.

count_ntm_stops

:   Count of transit stops reported to National Transit Map as of 1/2023

stops_per_capita

:   Transit stops per 1000 people

stops_per_sqmile

:   Transit stops per square mile

census_tract_area20

:   Census land area, square miles 