---
title: "Does public transportation decrease social isolation?"
author: "Naomi Buell and Richie Rivera"
format: html
bibliography: references.bib
echo: false
warning: false
---

## Load data

```{r}
#| label: load packages
#| echo: false
#| include: false

library(tidyverse)
library(janitor)
library(httr)
library(jsonlite)
library(skimr)
library(RSocrata)
library(CDCPLACES)
library(readr)
library(ggplot2)
library(corrplot)
library(tidycensus)
library(caret)
```

Load [PLACES: Local Data for Better Health, Census Tract Data 2024 release](https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-Census-Tract-D/cwsq-ngmh/about_data) using [@CDCPLACES] [@Greenlund2022]. This is 2022 data.

```{r}
#| label: get-cdc-data-with-CDCPLACES-package
dict <- get_dictionary()

cdc_places <- CDCPLACES::get_places(
    geography = "census",
    measure = c("ISOLATION"),
    age_adjust = TRUE
) |>
    clean_names() |>
    mutate(totalpopulation = as.numeric(totalpopulation)) |>
    select(
        stateabbr,
        locationname,
        data_value,
        totalpopulation
    ) # Consider pulling geolocation variable if mapping

# Browse
cdc_places |> head()
```

Get [Public Transit Stops by Census Tract and ZIP Code Tabulation Area, United States, 2024](https://www.icpsr.umich.edu/web/ICPSR/studies/38605/datadocumentation#) from the National Neighborhood Data Archive (NaNDA) [@pan2023].

```{r}
#| label: get-transit

transit <- read_table("data/ICPSR_38605/DS0003/38605-0003-Data.tsv") |>
    clean_names() |>
    mutate_all(~ replace(., is.na(.), 0)) # Filling the empty with 0 since they correspond with national parks

# Browse
transit |> head()
```

Get 2010 USDA-ERS Rural Urban Commuting Area (RUCA) Code classification scheme for Census tracts [@paykin2022].

```{r}
#| label: get-RUCA

ruca <- read_csv(
    "https://raw.githubusercontent.com/spaykin/rural-urban-classification/refs/heads/main/data_final/RuralSubUrban_T.csv"
) |>
    clean_names() |>
    rename(tract_fips20 = tract_fips)

ruca |> head()
```

Get age by census tract from the [2020 Census](https://www2.census.gov/programs-surveys/decennial/2020/data/demographic-and-housing-characteristics-file/age-group-visualization/tract-agegroup-2020.csv) [@census2023].

```{r}
#| label: get-age

# Get FIPS lookup by state and county
data(fips_codes)
fips_codes <- fips_codes |>
    rename(stateabbr = state)

# Get age data from Census
age <- read_csv(
    # From Census website
    # "https://www2.census.gov/programs-surveys/decennial/2020/data/demographic-and-housing-characteristics-file/age-group-visualization/tract-agegroup-2020.csv"
    # From GitHub repo
    "https://github.com/naomibuell/data-646-final-project/raw/refs/heads/main/data/tract-agegroup-2020.csv"
) |>
    clean_names() |>
    mutate(
        rate_65_plus = (x65_to_84_years_2020_population_percent + x85_to_99_years_2020_population_percent + x100_years_and_over_2020_population_percent) / 100,
        rate_85_plus = (x85_to_99_years_2020_population_percent + x100_years_and_over_2020_population_percent) / 100,
        totalpopulation = all_ages_2020_population_number |> as.numeric()
    ) |>
    rename(state_name = state) |>
    full_join(fips_codes) |>
    mutate(
        tract_num = str_extract(tract, "(?<=Census Tract )\\d+(\\.\\d+)?"),
        tract_padded = str_pad(str_replace(tract_num, "\\.", ""), 6, pad = "0", side = "right"),
        tract_fips20 = paste0(state_code, county_code, tract_padded)
    ) |>
    select(
        stateabbr,
        tract_fips20,
        contains("rate")
    )
# Browse
age |> head()
```

Merge data. Since there are two columns for total population, we will use the one that is more complete, `totalpopulation`, from the CDC PLACES data. We will also rename the columns for clarity and consistency and generate a new variable, `density_per_sqmi`, which is the total population divided by the area in square miles.

```{r}
#| label: merge-rename

data <- cdc_places |>
    rename(tract_fips20 = locationname) |>
    full_join(transit) |>
    full_join(ruca) |>
    full_join(age) |>
    rename(
        area_sq_miles = census_tract_area20, # Area in square miles from transit data
        fips = tract_fips20, # FIPS code for census tract
        isolation = data_value, # Renaming for clarity
        ruca = ruca2, # using RUCA code 2 only
        state_abbr = stateabbr, # Renaming for consistency
    ) |>
    mutate(
        state_abbr = state_abbr |> as.factor(), # Ensure state abbreviation is a factor
        fips = fips |> as.factor(), # Ensure FIPS code is a factor
        totalpopulation = as.numeric(totalpopulation), # Convert total population to numeric
        rurality = factor(rurality, levels = c("Urban", "Suburban", "Rural")), # Reorder rurality factor levels
        ruca = case_when(
            ruca == 99 ~ NA, # Replace RUCA code 99 (not coded) with NA
            TRUE ~ ruca
        ) |>
            factor(),
        density_per_sqmi = totalpopulation / area_sq_miles # creating a population land density measure from  population size (CDC PLACES) and area in square miles (transit data)
    ) |>
    filter(!is.na(isolation)) |> # Remove rows with missing isolation data
    select(-c(ruca1, tot_pop_2020)) # Remove unnecessary columns

# Preview
data |> head()
```

## Exploratory data analysis and data preparation

### Missingness

```{r}
#| label: skim
data |> skim()

perc_missing_transit <- sum(is.na(data$count_ntm_stops)) /
    nrow(data) *
    100 |> round()
perc_missing_ruca <- sum(is.na(data$ruca)) / nrow(data) * 100 |> round()
perc_missing_age <- sum(is.na(data$rate_65_plus)) / nrow(data) * 100 |> round()
```

After filtering raw data for non-missing observations of our dependent variable, feelings of social isolation (`isolation`), we have `r nrow(data)` observations. County FIPs and state columns are 100% complete. A small number of counties in the data have missing transit data (`r perc_missing_transit` % missing) and some counties are not categorized into urban/suburban/rural categories (`r perc_missing_ruca` % missing). The age data has significant missingness (`r perc_missing_age` % missing).

```{r}
#| label: perc-missing-by-state

data |>
    select(state_abbr, count_ntm_stops) |>
    group_by(state_abbr) |>
    skim()

all_states <- state.abb # Complete list of state abbreviations
present_states <- unique(data$state_abbr) # Unique, non-missing state abbreviations in your data
missing_states <- setdiff(all_states, present_states) # Identify missing states
```

Note that all counties in Connecticut (CT) have missing transit data, which is why we will remove it from our analysis. The state abbreviation "CT" has `r sum(data$state_abbr == "CT")` rows with missing transit data. The following states are completely missing for the dataset (no CDC PLACES isolation data available): `r paste(missing_states, collapse = ", ")`.

Revising data by filteringing out missing information (dropped observations are documented in our paper).

```{r}
#| label: remove-missings

df_nonmiss <- data |>
    filter(state_abbr != "CT") # Removing CT (all missing)

skim(df_nonmiss) # Skim the non-missing dataframe

# 11001980000 - NATIONAL MALL   - has a pop of 17
# 36061014300 - CENTRAL PARK    - has a pop of 1
```

Now there are `r nrow(df_nonmiss)` observations remaining after removing Connecticut, and there are no missing isolation or transit data. We note a few census tracts with very low populations (e.g., 1 person in central park or 17 people in the national mall) that may not be representative of the general population.

### Distribution

Viewing the distribution of our modeling variables.

```{r}
#| label: histograms
#| warning: false
df_hist <- df_nonmiss |>
    select(
        count_ntm_stops,
        stops_per_capita,
        density_per_sqmi,
        isolation
    ) |>
    pivot_longer(
        cols = everything(),
        names_to = "variable",
        values_to = "value"
    )

ggplot(df_hist, aes(x = value)) +
    geom_histogram(
        fill = "steelblue",
        color = "black"
    ) +
    facet_wrap(~variable, scales = "free") +
    theme_classic() +
    labs(
        title = "Distribution of Transit Stops, Population Density, and Isolation Measures",
        subtitle = "Transit and density measures are right-skewed. Isolation is normally distributed.",
        caption = "Source: CDC PLACES and National Neighborhood Data Archive (NaNDA)",
        x = "Value",
        y = "Count"
    ) +
    scale_x_continuous(labels = scales::comma) +
    scale_y_continuous(labels = scales::comma)

df_nonmiss |>
    select(stops_per_capita, density_per_sqmi, count_ntm_stops) |>
    skim()
```

The column `stops_per_capita` has high outliers (e.g., 12k stops per capita observation when mean value is 2). We plot the data with outliers (defined as `Q3 + 1.5 * IQR`) removed.\*\*

```{r}
#| label: remove-outliers
get_upper_outlier_threshold <- function(data) {
    Q3 <- quantile(data, 0.75) # Third quartile
    IQR_value <- IQR(data) # Interquartile range
    upper_bound <- Q3 + 1.5 * IQR_value
    return(upper_bound)
}

df_outliers <- df_nonmiss |>
    filter(
        stops_per_capita <
            get_upper_outlier_threshold(df_nonmiss$stops_per_capita)
    ) |>
    select(
        count_ntm_stops,
        stops_per_capita,
        density_per_sqmi,
        isolation
    ) |>
    pivot_longer(
        cols = everything(),
        names_to = "variable",
        values_to = "value"
    ) |>
    mutate(
        variable = recode(
            variable,
            count_ntm_stops = "Number of Stops",
            stops_per_capita = "Stops Per Capita",
            density_per_sqmi = "Population Density per Square Mile",
            isolation = "Isolation"
        )
    )

ggplot(
    df_outliers,
    aes(x = value, y = after_stat(count) / sum(after_stat(count)) * 100)
) +
    geom_histogram(
        fill = "steelblue",
        color = "black"
    ) +
    facet_wrap(~variable, scales = "free", ncol = 1) +
    theme_classic() +
    labs(x = "Value", y = "% of Total") +
    scale_x_continuous(labels = scales::comma)
```

We also investigate distributions of the centered and scaled data.

```{r}
#| label: histograms-scaled
df_hist <- df_nonmiss |>
    filter(
        stops_per_capita <
            get_upper_outlier_threshold(df_nonmiss$stops_per_capita)
    ) |>
    select(
        count_ntm_stops,
        stops_per_capita,
        density_per_sqmi,
        isolation
    ) |>
    mutate(across(everything(), scale)) |>
    pivot_longer(
        cols = everything(),
        names_to = "variable",
        values_to = "value"
    ) |>
    mutate(
        variable = recode(
            variable,
            count_ntm_stops = "Number of Stops",
            stops_per_capita = "Stops Per Capita",
            density_per_sqmi = "Density per Square Mile",
            isolation = "Isolation"
        )
    )

ggplot(
    df_hist,
    aes(x = value, y = after_stat(count) / sum(after_stat(count)) * 100)
) +
    geom_histogram(
        fill = "steelblue",
        color = "black"
    ) +
    facet_wrap(~variable, scales = "free", ncol = 1) +
    theme_classic() +
    labs(x = "Value", y = "% of Total")
```

### Correlation

The figure below shows that feelings of social isolation among adults is most correlated with population density, number of stops, stops per square mile, and rurality. Positive correlation indicates that the denser an area is, the higher the rate of feeling social isolation, which is counterintuitive to our hypothesis. Social isolation is also positively associated with the count of transit stops (total count in census tract and adjusted per square mile)--also counterintuitive. Social isolation is slightly negatively correlated with rurality as measured by RUCA code, where a higher prevalence of this feeling tends to be associated with a more urban, less rural area, which is also inconsistent with our expectation.

```{r}
#| label: corrplot

renamed_df <- df_nonmiss |>
    mutate("Rurality (RUCA code)" = as.numeric(ruca)) |>
    rename(
        "Number of Stops" = count_ntm_stops,
        "Stops per 1k Capita" = stops_per_capita,
        "Stops per Square Mile" = stops_per_sqmile,
        "Population Density" = density_per_sqmi,
        "Feelings of Social Isolation" = isolation,
        "RUCA" = ruca,
        "Census Tract Area (Square Miles)" = area_sq_miles,
        "Total Population" = totalpopulation,
        "Percent 65 and up" = rate_65_plus,
        "Percent 85 and up" = rate_85_plus
    )

cor_matrix <- renamed_df |>
    select_if(is.numeric) |>
    cor(use = "pairwise.complete.obs") # Compute correlation matrix, not including NAs (in pairs)

corrplot(
    cor_matrix,
    method = "circle",
    type = "upper",
    tl.srt = 20,
    # diag = FALSE,
    order = "hclust"
)
```

There are no columns with pair-wise correlations above the threshold .75, so we don't need to filter out any highly correlated predictor variables.

```{r}
#| label: find-correlations
#| echo: false
findCorrelation(cor(df_nonmiss |> select_if(is.numeric)), cutoff = .75)
```

## Analysis

First, we will fit a linear model to the data for simplicity.

```{r}
#| label: fit-a-linear-model
#| echo: false
#| warning: false

# Removing state_abbr, fips, and ruca from the predictors
df_predictors <- df_nonmiss |>
    select(-c(state_abbr, fips, ruca))

lm_model <- lm(
    isolation ~ .,
    data = df_predictors
)
summary(lm_model)
```

-   `totalpopulation`: For each additional person in a census tract, the probability who report feeling socially isolated **`r ifelse(lm_model$coefficients["totalpopulation"] < 0, "decreases", "increases")` by `r abs(round(lm_model$coefficients["totalpopulation"] * 100, 2))`%**.

-   `area_sq_miles`: For each additional square mile in a census tract, the probability who report feeling socially isolated **`r ifelse(lm_model$coefficients["area_sq_miles"] < 0, "decreases", "increases")` by `r abs(round(lm_model$coefficients["area_sq_miles"] * 100, 2))`%**.

-   `count_ntm_stops`: For each additional transit stop in a census tract, the probability who report feeling socially isolated **`r ifelse(lm_model$coefficients["count_ntm_stops"] < 0, "decreases", "increases")` by `r abs(round(lm_model$coefficients["count_ntm_stops"] * 100, 2))`%**.

-   `stops_per_sqmile`: For each additional stop per square mile, the probability who report feeling socially isolated in a census tract **`r ifelse(lm_model$coefficients["stops_per_sqmile"] < 0, "decreases", "increases")` by `r abs(round(lm_model$coefficients["stops_per_sqmile"] * 100, 2))`%**.

-   `ruralitySuburban`: If the census tract is considered suburban, the probability who report feeling socially isolated **`r ifelse(lm_model$coefficients["ruralitySuburban"] < 0, "decreases", "increases")` by `r abs(round(lm_model$coefficients["ruralitySuburban"] * 100, 2))`%** compared to an urban tract.

-   `ruralityRural`: If the census tract is considered rural, the probability who report feeling socially isolated **`r ifelse(lm_model$coefficients["ruralityRural"] < 0, "decreases", "increases")` by `r abs(round(lm_model$coefficients["ruralityRural"] * 100, 2))`%** compared to an urban tract.

-   `rate_65_plus`: For each additional percentage point of the population aged 65 and older, the probability who report feeling socially isolated in a tract **`r ifelse(lm_model$coefficients["rate_65_plus"] < 0, "decreases", "increases")` by `r abs(round(lm_model$coefficients["rate_65_plus"] * 100, 2))`%**.

-   `rate_85_plus`: For each additional percentage point of the population aged 85 and older, the probability who report feeling socially isolated in a tract **`r ifelse(lm_model$coefficients["rate_85_plus"] < 0, "decreases", "increases")` by `r abs(round(lm_model$coefficients["rate_85_plus"] * 100, 2))`%**.

-   `density_per_sqmi`: For each additional person per square mile, the probability who report feeling socially isolated in a tract **`r ifelse(lm_model$coefficients["density_per_sqmi"] < 0, "decreases", "increases")` by `r abs(round(lm_model$coefficients["density_per_sqmi"] * 100, 2))`%**.

-   `stops_per_capita` was **insignificant**, indicating that the number of transit stops per capita does not significantly affect feelings of social isolation in this model.

The model's R-squared value indicates that approximately `r round(summary(lm_model)$r.squared * 100, 2)`% of the variance in feelings of social isolation can be explained by the model.

Next, we try bucketing the data by RUCA code. We will use `rurality` that is a factor with three levels: urban, suburban, and rural.

```{r}
#| label: lm-rurality
```

## Data Dictionary

`isolation`

:   Probability among adults who report always/usually/sometimes feeling socially isolated. Source: CDC PLACES.

`totalpopulation`

:   Total population of Census 2020. Source: CDC PLACES.

`count_ntm_stops`

:   Count of transit stops reported to National Transit Map as of 1/2023. Source: NaNDA.

`stops_per_capita`

:   Transit stops per 1000 people. Source: NaNDA.

`stops_per_sqmile`

:   Transit stops per square mile. Source: NaNDA.

`area_sq_miles`

:   Census land area, square miles. Source: NaNDA.

`ruca`

:   Secondary RUCA codes, 2010. Source: 2010 USDA-ERS.

    | **Code** | **Classification description** |
    |-----------------------|-------------------------------------------------|
    | 1 Metropolitan area core: primary flow within an urbanized area (UA) |  |
    | 1.0 | No additional code |
    | 1.1 | Secondary flow 30% to 50% to a larger UA |
    | 2 Metropolitan area high commuting: primary flow 30% or more to a UA |  |
    | 2.0 | No additional code |
    | 2.1 | Secondary flow 30% to 50% to a larger UA |
    | 3 Metropolitan area low commuting: primary flow 10% to 30% to a UA |  |
    | 3.0 | No additional code |
    | 4 Micropolitan area core: primary flow within an urban cluster of 10,000 to 49,999 (large UC) |  |
    | 4.0 | No additional code |
    | 4.1 | Secondary flow 30% to 50% to a UA |
    | 5 Micropolitan high commuting: primary flow 30% or more to a large UC |  |
    | 5.0 | No additional code |
    | 5.1 | Secondary flow 30% to 50% to a UA |
    | 6 Micropolitan low commuting: primary flow 10% to 30% to a large UC |  |
    | 6.0 | No additional code |
    | 7 Small town core: primary flow within an urban cluster of 2,500 to 9,999 (small UC) |  |
    | 7.0 | No additional code |
    | 7.1 | Secondary flow 30% to 50% to a UA |
    | 7.2 | Secondary flow 30% to 50% to a large UC |
    | 8 Small town high commuting: primary flow 30% or more to a small UC |  |
    | 8.0 | No additional code |
    | 8.1 | Secondary flow 30% to 50% to a UA |
    | 8.2 | Secondary flow 30% to 50% to a large UC |
    | 9 Small town low commuting: primary flow 10% to 30% to a small UC |  |
    | 9.0 | No additional code |
    | 10 Rural areas: primary flow to a tract outside a UA or UC |  |
    | 10.0 | No additional code |
    | 10.1 | Secondary flow 30% to 50% to a UA |
    | 10.2 | Secondary flow 30% to 50% to a large UC |
    | 10.3 | Secondary flow 30% to 50% to a small UC |
    | 99 Not coded: Census tract has zero population and no rural-urban identifier information |  |