---
title: "Does public transportation decrease social isolation?"
author: "Naomi Buell and Richie Rivera"
format: html
bibliography: references.bib
echo: false
warning: false
---

## Load data

```{r}
#| label: load packages
#| echo: false
#| include: false

library(tidyverse)
library(janitor)
library(httr)
library(jsonlite)
library(skimr)
library(CDCPLACES)
library(readr)
library(ggplot2)
library(corrplot)
library(tidycensus)
library(caret)
library(broom)
```

Load [PLACES: Local Data for Better Health, Census Tract Data 2024 release](https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-Census-Tract-D/cwsq-ngmh/about_data) using [@CDCPLACES] [@Greenlund2022]. This is 2022 data.

```{r}
#| label: get-cdc-data-with-CDCPLACES-package
dict <- get_dictionary()

cdc_places <- CDCPLACES::get_places(
    geography = "census",
    measure = c("ISOLATION"),
    age_adjust = TRUE
) |>
    clean_names() |>
    mutate(
        totalpop18plus = as.numeric(totalpop18plus),
        totalpopulation = as.numeric(totalpopulation)
    ) |>
    select(
        stateabbr,
        locationname,
        data_value,
        totalpop18plus,
        totalpopulation
    ) # Consider pulling geolocation variable if mapping

# Browse
cdc_places |> head()
```

Get [Public Transit Stops by Census Tract and ZIP Code Tabulation Area, United States, 2024](https://www.icpsr.umich.edu/web/ICPSR/studies/38605/datadocumentation#) from the National Neighborhood Data Archive (NaNDA) [@pan2023].

```{r}
#| label: get-transit

transit <- read_table("data/ICPSR_38605/DS0003/38605-0003-Data.tsv") |>
    clean_names() |>
    mutate_all(~ replace(., is.na(.), 0)) # Filling the empty with 0 since they correspond with national parks

# Browse
transit |> head()
```

Get 2010 USDA-ERS Rural Urban Commuting Area (RUCA) Code classification scheme for Census tracts [@paykin2022].

```{r}
#| label: get-RUCA

ruca <- read_csv(
    "https://raw.githubusercontent.com/spaykin/rural-urban-classification/refs/heads/main/data_final/RuralSubUrban_T.csv"
) |>
    clean_names() |>
    rename(tract_fips20 = tract_fips)

ruca |> head()
```

Get age by census tract from the [2020 Census](https://www2.census.gov/programs-surveys/decennial/2020/data/demographic-and-housing-characteristics-file/age-group-visualization/tract-agegroup-2020.csv) [@census2023].

```{r}
#| label: get-age

# Get FIPS lookup by state and county
data(fips_codes)
fips_codes <- fips_codes |>
    rename(stateabbr = state)

# Get age data from Census
age <- read_csv(
    # From Census website
    # "https://www2.census.gov/programs-surveys/decennial/2020/data/demographic-and-housing-characteristics-file/age-group-visualization/tract-agegroup-2020.csv"
    # From GitHub repo
    "https://github.com/naomibuell/data-646-final-project/raw/refs/heads/main/data/tract-agegroup-2020.csv"
) |>
    clean_names() |>
    mutate(
        rate_65_plus = (x65_to_84_years_2020_population_percent + x85_to_99_years_2020_population_percent + x100_years_and_over_2020_population_percent),
        rate_85_plus = (x85_to_99_years_2020_population_percent + x100_years_and_over_2020_population_percent),
        totalpopulation = all_ages_2020_population_number |> as.numeric()
    ) |>
    rename(state_name = state) |>
    full_join(fips_codes) |>
    mutate(
        tract_num = str_extract(tract, "(?<=Census Tract )\\d+(\\.\\d+)?"),
        tract_padded = str_pad(str_replace(tract_num, "\\.", ""), 6, pad = "0", side = "right"),
        tract_fips20 = paste0(state_code, county_code, tract_padded)
    ) |>
    select(
        stateabbr,
        tract_fips20,
        contains("rate")
    )
# Browse
age |> head()
```

Merge data. Since there are two columns for total population, we will use the one that is more complete, `totalpopulation`, from the CDC PLACES data. We will also rename the columns for clarity and consistency and generate a new variable, `density_per_sqmi`, which is the total population divided by the area in square miles.

```{r}
#| label: merge-rename

data <- cdc_places |>
    rename(tract_fips20 = locationname) |>
    full_join(transit) |>
    full_join(ruca) |>
    full_join(age) |>
    rename(
        area_sq_miles = census_tract_area20, # Area in square miles from transit data
        fips = tract_fips20, # FIPS code for census tract
        isolation = data_value, # Renaming for clarity
        ruca = ruca2, # using RUCA code 2 only
        state_abbr = stateabbr, # Renaming for consistency
    ) |>
    mutate(
        state_abbr = state_abbr |> as.factor(), # Ensure state abbreviation is a factor
        fips = fips |> as.factor(), # Ensure FIPS code is a factor
        rurality = factor(rurality, levels = c("Urban", "Suburban", "Rural")), # Reorder rurality factor levels
        ruca = case_when(
            ruca == 99 ~ NA, # Replace RUCA code 99 (not coded) with NA
            TRUE ~ ruca
        ) |>
            factor(),
        density_per_sqmi = totalpop18plus / area_sq_miles # creating a population land density measure from  population size (CDC PLACES) and area in square miles (transit data)
    ) |>
    filter(!is.na(isolation)) |> # Remove rows with missing isolation data
    select(-c(ruca1, tot_pop_2020)) # Remove unnecessary columns

# Preview
data |> head()
```

## Exploratory data analysis and data preparation

### Missingness

```{r}
#| label: skim
data |> skim()

perc_missing_transit <- sum(is.na(data$count_ntm_stops)) /
    nrow(data) *
    100 |> round()
perc_missing_ruca <- sum(is.na(data$ruca)) / nrow(data) * 100 |> round()
perc_missing_age <- sum(is.na(data$rate_65_plus)) / nrow(data) * 100 |> round()
```

After filtering raw data for non-missing observations of our dependent variable, feelings of social isolation (`isolation`), we have `r nrow(data)` observations. County FIPs and state columns are 100% complete. A small number of counties in the data have missing transit data (`r perc_missing_transit` % missing) and some counties are not categorized into urban/suburban/rural categories (`r perc_missing_ruca` % missing). The age data has significant missingness (`r perc_missing_age` % missing).

```{r}
#| label: perc-missing-by-state

data |>
    select(state_abbr, count_ntm_stops) |>
    group_by(state_abbr) |>
    skim()

all_states <- state.abb # Complete list of state abbreviations
present_states <- unique(data$state_abbr) # Unique, non-missing state abbreviations in your data
missing_states <- setdiff(all_states, present_states) # Identify missing states
```

Note that all counties in Connecticut (CT) have missing transit data, which is why we will remove it from our analysis. The state abbreviation "CT" has `r sum(data$state_abbr == "CT")` rows with missing transit data. The following states are completely missing for the dataset (no CDC PLACES isolation data available): `r paste(missing_states, collapse = ", ")`.

Revising data by filtering out missing information (dropped observations are documented in our paper).

```{r}
#| label: remove-missings

df_nonmiss <- data |>
    filter(
        !is.na(stops_per_capita), # Remove rows with missing stops_per_capita
        !is.na(isolation), # Remove rows with missing isolation data
        !is.na(rate_65_plus), # Remove rows with missing rate_65_plus
        !is.na(ruca) # Remove rows with missing RUCA data
    )

skim(df_nonmiss) # Skim the non-missing dataframe

# 11001980000 - NATIONAL MALL   - has a pop of 17
# 36061014300 - CENTRAL PARK    - has a pop of 1
```

Now there are `r nrow(df_nonmiss)` observations remaining after removing Connecticut and observations with missing rates of the elderly population, and there are no missing isolation or transit data. We note a few census tracts with very low populations (e.g., 1 person in central park or 17 people in the national mall) that may not be representative of the general population.

### Distribution

Viewing the distribution of our modeling variables.

```{r}
#| label: histograms
#| warning: false
df_hist <- df_nonmiss |>
    select(
        #count_ntm_stops,
        #stops_per_capita,
        #density_per_sqmi,
        #isolation,
        rate_65_plus
    ) |>
    pivot_longer(
        cols = everything(),
        names_to = "variable",
        values_to = "value"
    )

ggplot(df_hist, aes(x = value)) +
    geom_histogram(
        fill = "steelblue",
        color = "black"
    ) +
    facet_wrap(~variable, scales = "free") +
    theme_classic() +
    labs(
        title = "Distribution of Transit Stops, Population Density, and Isolation Measures",
        subtitle = "Transit and density measures are right-skewed. Isolation is normally distributed.",
        caption = "Source: CDC PLACES and National Neighborhood Data Archive (NaNDA)",
        x = "Value",
        y = "Count"
    ) +
    scale_x_continuous(labels = scales::comma) +
    scale_y_continuous(labels = scales::comma)

df_nonmiss |>
    select(stops_per_capita, density_per_sqmi, count_ntm_stops) |>
    skim()
```

The column `stops_per_capita` has high outliers (e.g., 12k stops per capita observation when mean value is 2). We plot the data with outliers (defined as `Q3 + 1.5 * IQR`) removed.\*\*

```{r}
ggplot(df_nonmiss, aes(x = rurality, y = isolation)) +
  geom_boxplot() +
  labs(
    title = "Isolation by Rurality",
    x = "Rurality",
    y = "Isolation"
  ) +
  theme_minimal()
```

```{r}
#| label: remove-outliers
get_upper_outlier_threshold <- function(data) {
    Q3 <- quantile(data, 0.75) # Third quartile
    IQR_value <- IQR(data) # Interquartile range
    upper_bound <- Q3 + 1.5 * IQR_value
    return(upper_bound)
}

df_outliers <- df_nonmiss |>
    filter(
        stops_per_capita <
            get_upper_outlier_threshold(df_nonmiss$stops_per_capita)
    )

ggplot(
    df_outliers |>
        select(
            count_ntm_stops,
            stops_per_capita,
            density_per_sqmi,
            isolation
        ) |>
        pivot_longer(
            cols = everything(),
            names_to = "variable",
            values_to = "value"
        ) |>
        mutate(
            variable = recode(
                variable,
                count_ntm_stops = "Number of Stops",
                stops_per_capita = "Stops Per Capita",
                density_per_sqmi = "Population Density per Square Mile",
                isolation = "Isolation"
            )
        ),
    aes(x = value, y = after_stat(count) / sum(after_stat(count)) * 100)
) +
    geom_histogram(
        fill = "steelblue",
        color = "black"
    ) +
    facet_wrap(~variable, scales = "free", ncol = 2) +
    theme_classic() +
    labs(x = "Value", y = "% of Total") +
    scale_x_continuous(labels = scales::comma)
```

We also investigate distributions of the centered and scaled data.

```{r}
#| label: histograms-scaled
df_hist <- df_outliers |>
    select(
        count_ntm_stops,
        stops_per_capita,
        density_per_sqmi,
        isolation
    ) |>
    mutate(across(everything(), scale)) |>
    pivot_longer(
        cols = everything(),
        names_to = "variable",
        values_to = "value"
    ) |>
    mutate(
        variable = recode(
            variable,
            count_ntm_stops = "Number of Stops",
            stops_per_capita = "Stops Per Capita",
            density_per_sqmi = "Density per Square Mile",
            isolation = "Isolation"
        )
    )

ggplot(
    df_hist,
    aes(x = value, y = after_stat(count) / sum(after_stat(count)) * 100)
) +
    geom_histogram(
        fill = "steelblue",
        color = "black"
    ) +
    facet_wrap(~variable, scales = "free", ncol = 2) +
    theme_classic() +
    labs(x = "Value", y = "% of Total")
```

### Correlation

The figure below shows that feelings of social isolation among adults is most correlated with population density, number of stops, stops per square mile, and rurality. Positive correlation indicates that the denser an area is, the higher the rate of feeling social isolation, which is counterintuitive to our hypothesis. Social isolation is also positively associated with the count of transit stops (total count in census tract and adjusted per square mile)--also counterintuitive. Social isolation is slightly negatively correlated with rurality as measured by RUCA code, where a higher prevalence of this feeling tends to be associated with a more urban, less rural area, which is also inconsistent with our expectation.

```{r}
#| label: corrplot
#| fig.width: 8
#| fig.height: 8

renamed_df <- df_outliers |>
    mutate("Rurality (RUCA code)" = as.numeric(ruca)) |>
    rename(
        "Number of Stops" = count_ntm_stops,
        "Stops per 1k Capita" = stops_per_capita,
        "Stops per Square Mile" = stops_per_sqmile,
        "Population Density" = density_per_sqmi,
        "Feelings of Social Isolation" = isolation,
        "RUCA" = ruca,
        "Census Tract Area (Square Miles)" = area_sq_miles,
        "Total Population" = totalpopulation,
        "Total Adult Population (18+)" = totalpop18plus,
        "Percent 65 and up" = rate_65_plus,
        "Percent 85 and up" = rate_85_plus
    )

cor_matrix <- renamed_df |>
    select_if(is.numeric) |>
    cor(use = "pairwise.complete.obs")

corrplot(
    cor_matrix,
    method = "circle",
    tl.srt = 20,
    tl.cex = 0.7,
    order = "hclust"
)
```

## Analysis

We will fit a linear model to the data for simplicity. We bucket the data by `rurality` that is a factor with three levels: urban, suburban, and rural.

```{r}
#| label: fit-a-linear-model
#| echo: false
#| warning: false

# Removing state_abbr, fips, and ruca as they are not relevant predictors for the model
df_predictors <- df_outliers |>
    select(-c(
        rate_85_plus, # Just using rate_65_plus for now
        count_ntm_stops, # Using stops_per_capita instead
        stops_per_sqmile, # Using stops_per_capita instead
        totalpopulation, # Using adult population instead because this aligns with the denominator used for the rate of social isolation
        totalpop18plus, # Removing totalpop18plus as it is highly correlated with density_per_sqmi.
        area_sq_miles, # Already contained in density per square mile predictor, so it becomes insignificant when modelling.
        state_abbr, # Removing state abbreviation for modeling
        fips, # Removing FIPS code for modeling
        ruca
    )) # Removing RUCA code for modeling

rurality_levels <- c("Urban", "Suburban", "Rural")
model_coefficients <- list()
model_r_squared <- list()
unscaled_model_coefficients <- list()

for (rural_level in rurality_levels) {
    print(rural_level)
    filtered_df <- df_predictors |>
        filter(rurality == rural_level) |>
        select(-rurality) # Remove rurality column for modeling

    # Check if there are enough rows to run the model
    if (nrow(filtered_df) >= 3) {
        # Separate the outcome variable and predictors
        isolation_col <- filtered_df$isolation
        predictor_cols_unscaled <- filtered_df |> select(-isolation) # Keep unscaled for parameters

        # Store scaling parameters
        original_means <- colMeans(predictor_cols_unscaled)
        original_sds <- apply(predictor_cols_unscaled, 2, sd)

        # Center and scale the predictor variables
        scaled_predictors <- scale(predictor_cols_unscaled, center = TRUE, scale = TRUE)
        scaled_predictors_df <- as.data.frame(scaled_predictors)

        # Combine the scaled predictors with the outcome variable
        model_data <- cbind(isolation = isolation_col, scaled_predictors_df)
        model_data <- as.data.frame(model_data)

        lm_model <- lm(
            isolation ~ .,
            data = model_data
        )

        current_rurality <- as.character(rural_level)

        scaled_coef_df <- tidy(lm_model) |>
            select(term, estimate, p.value)

        model_coefficients[[current_rurality]] <- tidy(lm_model) |>
            select(term, estimate, p.value) |>
            rename(
                setNames("estimate", !!current_rurality),
                setNames("p.value", paste0(!!current_rurality, "_p"))
            )

        # Extract R-squared
        model_summary <- summary(lm_model)
        print(model_summary)
        model_r_squared[[current_rurality]] <- model_summary$r.squared

        # Unscale and uncenter coefficients
        unscaled_coefs <- numeric(nrow(scaled_coef_df))
        names(unscaled_coefs) <- scaled_coef_df$term
        unscaled_intercept <- scaled_coef_df$estimate[scaled_coef_df$term == "(Intercept)"]

        for (term in scaled_coef_df$term) {
            if (term != "(Intercept)") {
                original_var_name <- gsub("scaled_", "", term)
                if (original_var_name %in% names(original_sds)) {
                    unscaled_coefs[term] <- scaled_coef_df$estimate[scaled_coef_df$term == term] / original_sds[original_var_name]
                    unscaled_intercept <- unscaled_intercept - (unscaled_coefs[term] * original_means[original_var_name])
                } else {
                    unscaled_coefs[term] <- NA
                }
            } else {
                unscaled_coefs[term] <- unscaled_intercept
            }
        }

        unscaled_model_coefficients[[current_rurality]] <- enframe(unscaled_coefs, name = "term", value = paste0("unscaled_", current_rurality))
    } else {
        current_rurality <- as.character(rural_level)
        empty_df <- data.frame(term = character(0), estimate = numeric(0), p.value = numeric(0))
        model_coefficients[[current_rurality]] <- rename(
            empty_df,
            setNames("estimate", !!current_rurality),
            setNames("p.value", paste0(!!current_rurality, "_p"))
        )
        warning(paste("Not enough data points for rurality level:", rural_level, ". Skipping model. At least 3 rows are required."))
    }
}

# Merge the coefficient estimates and p-values into a single dataframe
compiled_coefficients <- model_coefficients[[1]]
for (i in 2:length(model_coefficients)) {
    compiled_coefficients <- compiled_coefficients |>
        full_join(model_coefficients[[i]], by = "term")
}

print(compiled_coefficients)

# Merge the unscaled coefficients into a single dataframe
compiled_unscaled_coefficients <- unscaled_model_coefficients[[1]]
for (i in 2:length(unscaled_model_coefficients)) {
    compiled_unscaled_coefficients <- compiled_unscaled_coefficients |>
        full_join(unscaled_model_coefficients[[i]], by = "term")
}

print("Unscaled and Uncentered Coefficients:")
print(compiled_unscaled_coefficients)
```

```{r}
#| label: get-counts-for-mrkdwn

count_rurality <- df_predictors |>
    group_by(rurality) |>
    summarize(
        count = n(),
        perc = n() / nrow(df_predictors) * 100 |> round(2)
    )
```

For the rural model:

-   n = `r count_rurality |> filter(rurality == "Rural") |> select(count)`

-   The model's R-squared value indicates that approximately **`r round(model_r_squared$Rural * 100, 2)`%** of the variance in feelings of social isolation can be explained by the model. This is relatively low, compared to the models for urban and suburban populations.

-   `stops_per_capita`: Controlling for age and population density, for each additional transit stop per 1,000 people in a census tract, the probability of adults feeling socially isolated **`r ifelse(compiled_unscaled_coefficients$unscaled_Rural[compiled_unscaled_coefficients$term == "stops_per_capita"] < 0, "decreases", "increases")` by `r abs(round(compiled_unscaled_coefficients$unscaled_Rural[compiled_unscaled_coefficients$term == "stops_per_capita"] * 100, 2))`%** (p = `r round(compiled_coefficients$Rural_p[compiled_coefficients$term == "stops_per_capita"], 3)`).

-   `rate_65_plus`: For each percentage point increase in population aged 65 and older, the probability of adults feeling socially isolated **`r ifelse(compiled_unscaled_coefficients$unscaled_Rural[compiled_unscaled_coefficients$term == "rate_65_plus"] < 0, "decreases", "increases")` by `r abs(round(compiled_unscaled_coefficients$unscaled_Rural[compiled_unscaled_coefficients$term == "rate_65_plus"]* 100, 2))`%** (p = `r round(compiled_coefficients$Rural_p[compiled_coefficients$term == "rate_65_plus"], 3)`).

-   `density_per_sqmi`: For each additional person per square mile, the probability of adults feeling socially isolated **`r ifelse(compiled_unscaled_coefficients$unscaled_Rural[compiled_unscaled_coefficients$term == "density_per_sqmi"] < 0, "decreases", "increases")` by `r abs(round(compiled_unscaled_coefficients$unscaled_Rural[compiled_unscaled_coefficients$term == "density_per_sqmi"] * 100, 2))`%** (p = `r round(compiled_coefficients$Rural_p[compiled_coefficients$term == "density_per_sqmi"], 3)`).

For the suburban model:

-   n = `r count_rurality |> filter(rurality == "Suburban") |> select(count)`

-   The model's R-squared value indicates that approximately `r round(model_r_squared$Suburban * 100, 2)`% of the variance in feelings of social isolation can be explained by the model.

-   `stops_per_capita`: Controlling for age and population density, for each additional transit stop per 1,000 people in a census tract, the probability who report feeling socially isolated **`r ifelse(compiled_unscaled_coefficients$unscaled_Suburban[compiled_unscaled_coefficients$term == "stops_per_capita"] < 0, "decreases", "increases")` by `r abs(round(compiled_unscaled_coefficients$unscaled_Suburban[compiled_unscaled_coefficients$term == "stops_per_capita"]* 100, 2))`%** (p = `r round(compiled_coefficients$Suburban_p[compiled_coefficients$term == "stops_per_capita"], 3)`).

-   `rate_65_plus`: For each percentage point increase in population aged 65 and older, the probability who report feeling socially isolated **`r ifelse(compiled_unscaled_coefficients$unscaled_Suburban[compiled_unscaled_coefficients$term == "rate_65_plus"] < 0, "decreases", "increases")` by `r abs(round(compiled_unscaled_coefficients$unscaled_Suburban[compiled_unscaled_coefficients$term == "rate_65_plus"]* 100, 2))`%** (p = `r round(compiled_coefficients$Suburban_p[compiled_coefficients$term == "rate_65_plus"], 3)`).

-   `density_per_sqmi`: For each additional person per square mile, the probability who report feeling socially isolated **`r ifelse(compiled_unscaled_coefficients$unscaled_Suburban[compiled_unscaled_coefficients$term == "density_per_sqmi"] < 0, "decreases", "increases")` by `r abs(round(compiled_unscaled_coefficients$unscaled_Suburban[compiled_unscaled_coefficients$term == "density_per_sqmi"]* 100, 2))`%** (p = `r round(compiled_coefficients$Suburban_p[compiled_coefficients$term == "density_per_sqmi"], 3)`).

For the urban model:

-   n = `r count_rurality |> filter(rurality == "Urban") |> select(count)`

-   The model's R-squared value indicates that approximately `r round(model_r_squared$Urban * 100, 2)`% of the variance in feelings of social isolation can be explained by the model.

-   `stops_per_capita`: Controlling for age and population density, for each additional transit stop per 1,000 people in a census tract, the probability who report feeling socially isolated **`r ifelse(compiled_unscaled_coefficients$unscaled_Urban[compiled_unscaled_coefficients$term == "stops_per_capita"] < 0, "decreases", "increases")` by `r abs(round(compiled_unscaled_coefficients$unscaled_Urban[compiled_unscaled_coefficients$term == "stops_per_capita"]* 100, 2))`%** (p = `r round(compiled_coefficients$Urban_p[compiled_coefficients$term == "stops_per_capita"], 3)`).

-   `rate_65_plus`: For each percentage point increase in population aged 65 and older, the probability who report feeling socially isolated **`r ifelse(compiled_unscaled_coefficients$unscaled_Urban[compiled_unscaled_coefficients$term == "rate_65_plus"] < 0, "decreases", "increases")` by `r abs(round(compiled_unscaled_coefficients$unscaled_Urban[compiled_unscaled_coefficients$term == "rate_65_plus"]* 100, 2))`%** (p = `r round(compiled_coefficients$Urban_p[compiled_coefficients$term == "rate_65_plus"], 3)`).

-   `density_per_sqmi`: For each additional person per square mile, the probability who report feeling socially isolated **`r ifelse(compiled_unscaled_coefficients$unscaled_Urban[compiled_unscaled_coefficients$term == "density_per_sqmi"] < 0, "decreases", "increases")` by `r abs(round(compiled_unscaled_coefficients$unscaled_Urban[compiled_unscaled_coefficients$term == "density_per_sqmi"]* 100, 2))`%** (p = `r round(compiled_coefficients$Urban_p[compiled_coefficients$term == "density_per_sqmi"], 3)`).

## Conclusion

Our analysis suggests that census tracts with greater access to public transportation, measured by the number of transit stops per capita, tend to report slightly *higher* rates of feelings of social isolation among adults. This is counter to our hypothesis that tracts with more transit would have less social isolation.

Stratifying results by urbanicity and controlling for the proportion of older adults ages 65 and up, as well as population density, transit had the largest effect in suburban areas, increasing rates of social isolation by 57%, and the smallest effect in rural areas.... \[insert\].

Notably, missingness in rurality and age-related variables limits our ability to fully account for context and demographic confounders. The high proportion of missing values, especially for older adults (65+ and 85+), is a key limitation, given that social isolation disproportionately affects this group. Future analyses could consider imputation or inclusion of alternative demographic datasets to fill these gaps.

## Data Dictionary

`isolation`

:   The estimated age-adjusted prevalence of feeling socially isolated among adults aged \>= 18 years (%) in 2022. Probability among adults who report always/usually/sometimes feeling socially isolated. Source: CDC PLACES.

`totalpopulation`

:   Total population of Census 2020. Source: CDC PLACES.

`totalpop18plus`

:   Total population of adults aged 18 years and older in Census 2020. Source: CDC PLACES.

`rate_65_plus`

:   Percentage of the population aged 65 years and older in Census 2020. Source: 2020 Census.

`rate_85_plus`

:   Percentage of the population aged 85 years and older in Census 2020. Source: 2020 Census.

`count_ntm_stops`

:   Count of transit stops reported to National Transit Map as of 1/2023. Source: NaNDA.

`stops_per_capita`

:   Transit stops per 1000 people. Source: NaNDA.

`stops_per_sqmile`

:   Transit stops per square mile. Source: NaNDA.

`area_sq_miles`

:   Census land area, square miles. Source: NaNDA.

`rurality`

:   Rural, suburban, or urban classification of the census tract based on the 2010 USDA-ERS Rural-Urban Continuum Codes (RUCC) and Rural-Urban Commuting Area (RUCA) codes. Source: 2010 USDA-ERS.

`ruca`

:   Secondary RUCA codes, 2010. Source: 2010 USDA-ERS.

    | **Code** | **Classification description** |
    |-----------------------------------|------------------------------------------------|
    | 1 Metropolitan area core: primary flow within an urbanized area (UA) |  |
    | 1.0 | No additional code |
    | 1.1 | Secondary flow 30% to 50% to a larger UA |
    | 2 Metropolitan area high commuting: primary flow 30% or more to a UA |  |
    | 2.0 | No additional code |
    | 2.1 | Secondary flow 30% to 50% to a larger UA |
    | 3 Metropolitan area low commuting: primary flow 10% to 30% to a UA |  |
    | 3.0 | No additional code |
    | 4 Micropolitan area core: primary flow within an urban cluster of 10,000 to 49,999 (large UC) |  |
    | 4.0 | No additional code |
    | 4.1 | Secondary flow 30% to 50% to a UA |
    | 5 Micropolitan high commuting: primary flow 30% or more to a large UC |  |
    | 5.0 | No additional code |
    | 5.1 | Secondary flow 30% to 50% to a UA |
    | 6 Micropolitan low commuting: primary flow 10% to 30% to a large UC |  |
    | 6.0 | No additional code |
    | 7 Small town core: primary flow within an urban cluster of 2,500 to 9,999 (small UC) |  |
    | 7.0 | No additional code |
    | 7.1 | Secondary flow 30% to 50% to a UA |
    | 7.2 | Secondary flow 30% to 50% to a large UC |
    | 8 Small town high commuting: primary flow 30% or more to a small UC |  |
    | 8.0 | No additional code |
    | 8.1 | Secondary flow 30% to 50% to a UA |
    | 8.2 | Secondary flow 30% to 50% to a large UC |
    | 9 Small town low commuting: primary flow 10% to 30% to a small UC |  |
    | 9.0 | No additional code |
    | 10 Rural areas: primary flow to a tract outside a UA or UC |  |
    | 10.0 | No additional code |
    | 10.1 | Secondary flow 30% to 50% to a UA |
    | 10.2 | Secondary flow 30% to 50% to a large UC |
    | 10.3 | Secondary flow 30% to 50% to a small UC |
    | 99 Not coded: Census tract has zero population and no rural-urban identifier information |  |