---
title: "Does public transportation decrease social isolation?"
author: "Naomi Buell and Richie Rivera"
format: html
bibliography: references.bib
echo: false
warning: false
---

## Load data

```{r}
#| label: load packages
#| echo: false
#| include: false

library(tidyverse)
library(janitor)
library(httr)
library(jsonlite)
library(skimr)
library(RSocrata)
library(CDCPLACES)
library(readr)
library(ggplot2)
library(corrplot)
```

Load [PLACES: Local Data for Better Health, Census Tract Data 2024 release](https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-Census-Tract-D/cwsq-ngmh/about_data) using [@CDCPLACES] [@Greenlund2022]. This is 2022 data.

```{r}
#| label: get-cdc-data-with-CDCPLACES-package
dict <- get_dictionary()

cdc_places <- CDCPLACES::get_places(
    geography = "census",
    measure = c("ISOLATION"),
    # "LACKTRPT", "EMOTIONSPT"), // Removing lack of transport and emotional support variables.
    age_adjust = TRUE
) |>
    clean_names() |>
    select(
        stateabbr,
        locationname,
        data_value,
        starts_with("totalpopulation")
    ) # Consider pulling geolocation variable if mapping

# Browse
cdc_places |> head()
```

Get [Public Transit Stops by Census Tract and ZIP Code Tabulation Area, United States, 2024](https://www.icpsr.umich.edu/web/ICPSR/studies/38605/datadocumentation#) from the National Neighborhood Data Archive (NaNDA) [@pan2023].

```{r}
#| label: get-transit

transit <- read_table("data/ICPSR_38605/DS0003/38605-0003-Data.tsv") |>
    clean_names() |>
    mutate_all(~ replace(., is.na(.), 0)) # Filling the empty with 0 since they correspond with national parks

# Browse
transit |> head()
```

Get 2010 USDA-ERS Rural Urban Commuting Area (RUCA) Code classification scheme for Census tracts [@paykin2022].

```{r}
#| label: get-RUCA

ruca <- read_csv(
    "https://raw.githubusercontent.com/spaykin/rural-urban-classification/refs/heads/main/data_final/RuralSubUrban_T.csv"
) |>
    clean_names() |>
    rename(tract_fips20 = tract_fips) |>
    mutate(rurality = rurality |> as_factor())

ruca |> head()
```

Merge data.

```{r}
#| label: merge-rename

data <- cdc_places |>
    rename(tract_fips20 = locationname) |>
    full_join(transit) |>
    full_join(ruca) |>
    rename(
        tot_pop_2022 = totalpopulation, # Total population from CDC PLACES
        area_sq_miles = census_tract_area20, # Area in square miles from transit data
        fips = tract_fips20, # FIPS code for census tract
        isolation = data_value, # Renaming for clarity
        ruca = ruca2, # using RUCA code 2 only
        state_abbr = stateabbr, # Renaming for consistency
    ) |>
    mutate(
        state_abbr = state_abbr |> as.factor(), # Ensure state abbreviation is a factor
        fips = fips |> as.factor(), # Ensure FIPS code is a factor
        tot_pop_2022 = as.numeric(tot_pop_2022), # Convert total population to numeric
        rurality = factor(rurality, levels = c("Urban", "Suburban", "Rural")), # Reorder rurality factor levels
        ruca = factor(ruca) # Order RUCA codes numerically
    ) |>
    filter(!is.na(isolation)) |> # Remove rows with missing isolation data
    select(-ruca1) # Remove unnecessary columns

# Preview
data |> head()
```

## Exploratory data analysis and data preparation

### Missingness

```{r}
#| label: skim
data |> skim()

perc_missing_transit <- sum(is.na(data$count_ntm_stops)) /
    nrow(data) *
    100 |> round()
perc_missing_ruca <- sum(is.na(data$ruca)) / nrow(data) * 100 |> round()
```

After filtering raw data for non-missing observations of our dependent variable, feelings of social isolation (`isolation`), we have `r nrow(data)` observations. County FIPs and state columns are 100% complete. A small number of counties in the data have missing transit data (`r perc_missing_transit` % missing) and some counties are not categorized into urban/suburban/rural categories (`r perc_missing_ruca` % missing).

```{r}
#| label: perc-missing-by-state

data |>
    select(state_abbr, count_ntm_stops) |>
    group_by(state_abbr) |>
    skim()

all_states <- state.abb # Complete list of state abbreviations
present_states <- unique(data$state_abbr) # Unique, non-missing state abbreviations in your data
missing_states <- setdiff(all_states, present_states) # Identify missing states
```

Note that all counties in Connecticut (CT) have missing transit data, which is why we will remove it from our analysis. The state abbreviation "CT" has `r sum(data$state_abbr == "CT")` rows with missing transit data. The following states are completely missing for the dataset (no CDC PLACES isolation data available): `r paste(missing_states, collapse = ", ")`.

Revising data by filteringing out missing information (dropped observations are documented in our paper).

```{r}
#| label: remove-missings

df_nonmiss <- data |>
    filter(state_abbr != "CT") # Removing CT (all missing)

skim(df_nonmiss) # Skim the non-missing dataframe

# 11001980000 - NATIONAL MALL   - has a pop of 17
# 36061014300 - CENTRAL PARK    - has a pop of 1
```

Now there are `r nrow(df_nonmiss)` observations remaining after removing Connecticut. We will use this non-missing data for our analysis.

We also create a population density variable, `density_per_sqmi`, which is the total population divided by the area in square miles. This will help us understand how population density relates to feelings of social isolation.

```{r}
        density_per_sqmi = tot_pop_2022 / area_sq_miles, # creating a population land density measure from latest population size (CDC PLACES, 2022) and area in square miles (transit data)

```

### Distribution

Viewing the distribution of our modeling variables.

```{r}

hist_df <- df |>
    select(
        count_ntm_stops,
        stops_per_capita,
        density_per_sqmi,
        data_value
    ) |>
    pivot_longer(
        cols = everything(),
        names_to = "variable",
        values_to = "value"
    )

ggplot(hist_df, aes(x = value)) +
    geom_histogram(
        bins = 30,
        fill = "steelblue",
        color = "black",
        alpha = 0.7
    ) +
    facet_wrap(~variable, scales = "free") +
    theme_minimal() +
    labs(
        title = "Histograms of Stops Per Capita, Density Per Sqmi, and Data Value",
        x = "Value",
        y = "Count"
    )

df |>
    arrange(desc(stops_per_capita)) |>
    head(50)
```

The field `stops_per_capita` has high outliers (e.g., 12k stops per capita observation when mean value is 2). We remove outliers defined as `Q3 + 1.5 * IQR`.

```{r}
get_upper_outlier_threshold <- function(data) {
    Q3 <- quantile(data, 0.75) # Third quartile
    IQR_value <- IQR(data) # Interquartile range
    upper_bound <- Q3 + 1.5 * IQR_value

    return(upper_bound)
}

hist_df <- df |>
    filter(
        stops_per_capita < get_upper_outlier_threshold(df$stops_per_capita)
    ) |>
    select(
        count_ntm_stops,
        stops_per_capita,
        density_per_sqmi,
        data_value
    ) |>
    pivot_longer(
        cols = everything(),
        names_to = "variable",
        values_to = "value"
    ) |>
    mutate(
        variable = recode(
            variable,
            count_ntm_stops = "NTM Stops Count",
            stops_per_capita = "Stops Per Capita",
            density_per_sqmi = "Density per Sqmi",
            data_value = "Lonlieness"
        )
    )

ggplot(
    hist_df,
    aes(x = value, y = after_stat(count) / sum(after_stat(count)))
) +
    geom_histogram(
        bins = 30,
        fill = "steelblue",
        color = "black",
        alpha = 0.7
    ) +
    facet_wrap(~variable, scales = "free", ncol = 1) +
    theme_minimal() +
    labs(x = "", y = "% of Total")

df |>
    arrange(desc(stops_per_capita)) |>
    head(50)
```

Centering and scaling the data, we get distributions that look like this:

```{r}
get_upper_outlier_threshold <- function(data) {
    Q3 <- quantile(data, 0.75) # Third quartile
    IQR_value <- IQR(data) # Interquartile range
    upper_bound <- Q3 + 1.5 * IQR_value

    return(upper_bound)
}

hist_df <- df |>
    filter(
        stops_per_capita < get_upper_outlier_threshold(df$stops_per_capita)
    ) |>
    select(
        count_ntm_stops,
        stops_per_capita,
        density_per_sqmi,
        data_value
    ) |>
    mutate(across(everything(), scale)) |>
    pivot_longer(
        cols = everything(),
        names_to = "variable",
        values_to = "value"
    ) |>
    mutate(
        variable = recode(
            variable,
            count_ntm_stops = "NTM Stops Count",
            stops_per_capita = "Stops Per Capita",
            density_per_sqmi = "Density per Sqmi",
            data_value = "Lonlieness"
        )
    )

ggplot(
    hist_df,
    aes(x = value, y = after_stat(count) / sum(after_stat(count)))
) +
    geom_histogram(
        bins = 30,
        fill = "steelblue",
        color = "black",
        alpha = 0.7
    ) +
    facet_wrap(~variable, scales = "free", ncol = 1) +
    theme_minimal() +
    labs(x = "Value", y = "% of Total")

df |>
    arrange(desc(stops_per_capita)) |>
    head(50)
```

### Correlation

Figure 1 below shows that feelings of social isolation among adults is most correlated with population; the positive correlation indicates that the denser an area is, the higher the rate of feeling social isolation, which seems counterintuitive. Social isolation is also positively associated with the count of transit stops. Lastly, social isolation is slightly negatively correlated with RUCA code, where a higher prevalence of this feeling tends to be associated with a more metropolitan area.

```{r}
#| label: corrplot

renamed_df <- df
names(renamed_df) <- c(
    "state_abbr",
    "fips",
    "Number of Stops",
    "Stops per Capita",
    "Population Density",
    "Feelings of Social Isolation",
    "ruca1",
    "RUCA",
    "rurality",
    "Total Population",
    "Census Tract Area"
)

df_num <- renamed_df |> select(-c(fips, rurality, state_abbr, ruca1)) # Drop non-numeric and duplicative variables
cor_matrix <- cor(df_num, use = "pairwise.complete.obs") # Compute correlation matrix, not including NAs (in pairs)
corrplot(
    cor_matrix,
    method = "circle",
    type = "upper",
    tl.srt = 20,
    # diag = F,
    order = "hclust"
)
```

## Analysis

```{r}
#| label: hypothesis-test
#| echo: false
#| warning: false

# Fit a linear model
lm_model <- lm(
    data_value ~
        count_ntm_stops + census_tract_area20 + tot_pop_2020 + rurality,
    data = df
)
summary(lm_model)
```

```{r}

lm_model <- lm(
    data_value ~ stops_per_capita + density_per_sqmi + ruca1,
    data = df
)
summary(lm_model)
```

The overall model is statistically significant (F-statistic p-value \< 2.2e-16). All variables are highly significant (p \< 0.001, indicated by \*\*\*):

-   stops_per_capita: For each unit increase, the data value increases by 0.001377

-   density_per_sqmi: For each unit increase, the data value increases by 0.0001861

-   ruca1: For each unit increase, the data value decreases by 0.04216

R-squared: 0.09819 (≈ 9.82%) This means the model explains about 9.82% of the variance in the data This is a relatively low R-squared, suggesting other important factors might not be included. The low R-squared suggests this model might not be the best fit for prediction.

### Hypothesis testing

## Data Dictionary

Feelings of social isolation among adults (`isolation`)

:   Probability among adults who report always/usually/sometimes feeling socially isolated

count_ntm_stops

:   Count of transit stops reported to National Transit Map as of 1/2023

stops_per_capita

:   Transit stops per 1000 people

stops_per_sqmile

:   Transit stops per square mile

census_tract_area20

:   Census land area, square miles

RUCA1

:   Primary RUCA Code

    | **Code** | **Classification description** |
    |--------------------|----------------------------------------------------|
    | 1 | Metropolitan area core: primary flow within an urbanized area (UA) |
    | 2 | Metropolitan area high commuting: primary flow 30% or more to a UA |
    | 3 | Metropolitan area low commuting: primary flow 10% to 30% to a UA |
    | 4 | Micropolitan area core: primary flow within an urban cluster of 10,000 to 49,999 (large UC) |
    | 5 | Micropolitan high commuting: primary flow 30% or more to a large UC |
    | 6 | Micropolitan low commuting: primary flow 10% to 30% to a large UC |
    | 7 | Small town core: primary flow within an urban cluster of 2,500 to 9,999 (small UC) |
    | 8 | Small town high commuting: primary flow 30% or more to a small UC |
    | 9 | Small town low commuting: primary flow 10% to 30% to a small UC |
    | 10 | Rural areas: primary flow to a tract outside a UA or UC |
    | 99 | Not coded: Census tract has zero population and no rural-urban identifier information |

RUCA2

:   Secondary RUCA codes, 2010

    | **Code** | **Classification description** |
    |-------------------|-----------------------------------------------------|
    | 1 Metropolitan area core: primary flow within an urbanized area (UA) |  |
    | 1.0 | No additional code |
    | 1.1 | Secondary flow 30% to 50% to a larger UA |
    | 2 Metropolitan area high commuting: primary flow 30% or more to a UA |  |
    | 2.0 | No additional code |
    | 2.1 | Secondary flow 30% to 50% to a larger UA |
    | 3 Metropolitan area low commuting: primary flow 10% to 30% to a UA |  |
    | 3.0 | No additional code |
    | 4 Micropolitan area core: primary flow within an urban cluster of 10,000 to 49,999 (large UC) |  |
    | 4.0 | No additional code |
    | 4.1 | Secondary flow 30% to 50% to a UA |
    | 5 Micropolitan high commuting: primary flow 30% or more to a large UC |  |
    | 5.0 | No additional code |
    | 5.1 | Secondary flow 30% to 50% to a UA |
    | 6 Micropolitan low commuting: primary flow 10% to 30% to a large UC |  |
    | 6.0 | No additional code |
    | 7 Small town core: primary flow within an urban cluster of 2,500 to 9,999 (small UC) |  |
    | 7.0 | No additional code |
    | 7.1 | Secondary flow 30% to 50% to a UA |
    | 7.2 | Secondary flow 30% to 50% to a large UC |
    | 8 Small town high commuting: primary flow 30% or more to a small UC |  |
    | 8.0 | No additional code |
    | 8.1 | Secondary flow 30% to 50% to a UA |
    | 8.2 | Secondary flow 30% to 50% to a large UC |
    | 9 Small town low commuting: primary flow 10% to 30% to a small UC |  |
    | 9.0 | No additional code |
    | 10 Rural areas: primary flow to a tract outside a UA or UC |  |
    | 10.0 | No additional code |
    | 10.1 | Secondary flow 30% to 50% to a UA |
    | 10.2 | Secondary flow 30% to 50% to a large UC |
    | 10.3 | Secondary flow 30% to 50% to a small UC |
    | 99 Not coded: Census tract has zero population and no rural-urban identifier information |  |