---
title: "Does public transportation decrease loneliness and social isolation?"
author: "Naomi Buell and Richie Rivera"
format: html
bibliography: references.bib
echo: false
warning: false
---

## Load data

```{r}
#| label: load packages
#| echo: false
#| include: false

library(tidyverse)
library(janitor)
library(httr)
library(jsonlite)
library(skimr)
library(RSocrata)
library(CDCPLACES)
library(readr)
library(ggplot2)
library(corrplot)
```

Load [PLACES: Local Data for Better Health, Census Tract Data 2024 release](https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-Census-Tract-D/cwsq-ngmh/about_data) using [@CDCPLACES] [@Greenlund2022]. This is 2022 data.

```{r}
#| label: get-cdc-data-with-CDCPLACES-package
dict <- get_dictionary()

cdc_places <- CDCPLACES::get_places(
  geography = "census",
  measure = c("ISOLATION"),
  # "LACKTRPT", "EMOTIONSPT"), // Removing lack of transport and emotional support variables.
  age_adjust = TRUE
) |>
  clean_names() |>
  select(stateabbr,
         locationname,
         data_value,
         starts_with("totalpopulation")) # Consider pulling geolocation variable if mapping

# Browse
cdc_places |> head()
```

Get [Public Transit Stops by Census Tract and ZIP Code Tabulation Area, United States, 2024](https://www.icpsr.umich.edu/web/ICPSR/studies/38605/datadocumentation#) from the National Neighborhood Data Archive (NaNDA) [@pan2023].

```{r}
#| label: get-transit

transit <- read_table("data/ICPSR_38605/DS0003/38605-0003-Data.tsv") |>
  clean_names() |>
  mutate_all( ~ replace(., is.na(.), 0)) # Filling the empty with 0 since they correspond with national parks

# Browse
transit |> head()
```

Get 2010 USDA-ERS Rural Urban Commuting Area (RUCA) Code classification scheme for Census tracts [@paykin2022].

```{r}
#| label: get-RUCA

ruca <- read_csv(
  "https://raw.githubusercontent.com/spaykin/rural-urban-classification/refs/heads/main/data_final/RuralSubUrban_T.csv"
) |>
  clean_names() |>
  rename(tract_fips20 = tract_fips) |>
  mutate(rurality = rurality |> as_factor())

ruca |> head()
```

Merge data.

```{r}
#| label: merge

data <- cdc_places |>
    rename(tract_fips20 = locationname) |>
    full_join(transit) |>
    full_join(ruca)

data |> head()
data |> skim()
```

## Exploratory data analysis

### Missingness

Investigate missings in merged data.

```{r}
#| label: browse-merged-data

nulls_df <- data |>
  select(stateabbr, count_ntm_stops) |>
  group_by(stateabbr) |>
  summarise(Null_Count = sum(is.na(count_ntm_stops)),
            Non_Null_Count = sum(!is.na(count_ntm_stops))) |>
  arrange(desc(Null_Count))

print(nulls_df, n = 100)
```

Investigate states with missing data.

```{r}
# Complete list of state abbreviations
all_states <- state.abb

# Unique, non-missing state abbreviations in your data
present_states <- unique(nulls_df$stateabbr)

# Determine missing states
missing_states <- setdiff(all_states, present_states)

# Display missing states
print(missing_states)
```

Filtering missing information (dropped observations are documented in our paper). CT is omitted because there is no data on it from the transit stop dataset.

```{r}
#| label: remove-missings

df <- data |>
  mutate(density_per_sqmi = tot_pop_2020 / census_tract_area20) |> # creating a population land density measure
  filter(stateabbr != "CT") |> # Removing CT
  filter(!is.na(stateabbr)) |> # Removing states where there is no data
  select(
    stateabbr,
    tract_fips20,
    count_ntm_stops,
    stops_per_capita,
    density_per_sqmi,
    data_value,
    ruca1,
    ruca2,
    rurality,
    tot_pop_2020,
    census_tract_area20
  ) # Selecting our columns of interest

head(df)

# 11001980000 - NATIONAL MALL   - has a pop of 17
# 36061014300 - CENTRAL PARK    - has a pop of 1
```

### Distribution

Viewing the distribution of our modeling variables.

```{r}

hist_df <- df |>
  select(count_ntm_stops,
         stops_per_capita,
         density_per_sqmi,
         data_value) |>
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value")

ggplot(hist_df, aes(x = value)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "black",
    alpha = 0.7
  ) +
  facet_wrap( ~ variable, scales = "free") +
  theme_minimal() +
  labs(title = "Histograms of Stops Per Capita, Density Per Sqmi, and Data Value", x = "Value", y = "Count")

df |>
  arrange(desc(stops_per_capita)) |>
  head(50)
```

The field `stops_per_capita` has high outliers (e.g., 12k stops per capita observation when mean value is 2). We remove outliers defined as `Q3 + 1.5 * IQR`.

```{r}
get_upper_outlier_threshold <- function(data) {
  Q3 <- quantile(data, 0.75) # Third quartile
  IQR_value <- IQR(data) # Interquartile range
  upper_bound <- Q3 + 1.5 * IQR_value
  
  return(upper_bound)
}

hist_df <- df |>
  filter(stops_per_capita < get_upper_outlier_threshold(df$stops_per_capita)) |>
  select(count_ntm_stops,
         stops_per_capita,
         density_per_sqmi,
         data_value) |>
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") |>
  mutate(
    variable = recode(variable, 
    count_ntm_stops = "NTM Stops Count", 
    stops_per_capita = "Stops Per Capita", 
    density_per_sqmi = "Density per Sqmi", 
    data_value = "Lonlieness")
    )

ggplot(hist_df, aes(x = value, y = after_stat(count) / sum(after_stat(count)))) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "black",
    alpha = 0.7
  ) +
  facet_wrap( ~ variable, scales = "free", ncol=1) +
  theme_minimal() +
  labs(x = "", y = "% of Total")

df |>
  arrange(desc(stops_per_capita)) |>
  head(50)
```

Centering and scaling the data, we get distributions that look like this:

```{r}
get_upper_outlier_threshold <- function(data) {
  Q3 <- quantile(data, 0.75) # Third quartile
  IQR_value <- IQR(data) # Interquartile range
  upper_bound <- Q3 + 1.5 * IQR_value
  
  return(upper_bound)
}

hist_df <- df |>
  filter(stops_per_capita < get_upper_outlier_threshold(df$stops_per_capita)) |>
  select(count_ntm_stops,
         stops_per_capita,
         density_per_sqmi,
         data_value) |>
  mutate(across(everything(), scale)) |>
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") |>
  mutate(
    variable = recode(variable, 
    count_ntm_stops = "NTM Stops Count", 
    stops_per_capita = "Stops Per Capita", 
    density_per_sqmi = "Density per Sqmi", 
    data_value = "Lonlieness")
    )

ggplot(hist_df, aes(x = value, y = after_stat(count) / sum(after_stat(count)))) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "black",
    alpha = 0.7
  ) +
  facet_wrap( ~ variable, scales = "free", ncol=1) +
  theme_minimal() +
  labs(x = "Value", y = "% of Total")

df |>
  arrange(desc(stops_per_capita)) |>
  head(50)
```

### Correlation

Figure 1 below shows that feelings of social isolation among adults (named `data_value`) is most correlated with population (`density_per_sqmi`); the positive correlation indicates that the denser an area is, the higher the rate of feeling social isolation, which seems counterintuitive. Social isolation is also positively associated with the count of transit stops (`count_ntm_stops`). Lastly, social isolation is slightly negatively correlated with RUCA code (`ruca2`), where a higher prevalence of this feeling tends to be associated with a more metropolitan area.

```{r}
#| label: corrplot

renamed_df <- df
names(renamed_df) <- c('stateabbr','tract_fips20','Number of Stops','Stops per Capita','Population Density','Reported Isolation','ruca1','RUCA','rurality','Total Population','Census Tract Area') 

df_num <- renamed_df |> select(-c(tract_fips20, rurality, stateabbr, ruca1))  # Drop non-numeric and duplicative variables 
cor_matrix <- cor(df_num, use = "pairwise.complete.obs")  # Compute correlation matrix, not including NAs (in pairs)
corrplot(cor_matrix,
         method = "circle",
         type = "upper",
         tl.srt = 20,
         #diag = F,
         order = "hclust")
```

## Hypothesis test

## Data Dictionary

Feelings of social isolation among adults

:   probability among adults who report always/usually/sometimes feeling socially isolated

:   Variable summary: Loneliness is defined as the discrepancy between a person’s desired and actual social relationships and is sometimes considered synonymous with social isolation, although they are two distinct concepts. Loneliness is an emotional response to social isolation, while social isolation is an objective measure of the lack of social interactions and relationships. Previous studies show that social isolation and loneliness are common sources of chronic stress in adults, and are common at the end of life, affecting 1 in 4 older adults. Positive relationships and interactions with family, friends, co-workers, and community members can have a protective impact on individual health and well-being, and these relationships can also help mitigate the negative impacts of challenges that people face (e.g., living in an unsafe neighborhood, trouble affording housing or food). Past research shows that people with high levels of social isolation are at higher risk of developing chronic conditions (e.g., depression, cardiovascular disease, hypertension, cancer) and have a higher risk of mortality.

Lack of reliable transportation in the past 12 months among adults

:   Nearly 2% of the U.S. population delayed medical care because they did not have transportation in 2017. Lack of available, convenient, or reliable transportation can affect a person’s ability to consistently access health care services, which can lead to delays in healthcare and medication use that can subsequently impact overall health. In addition, the inability to access reliable transportation to work, schools, and grocery stores is associated with higher rates of unemployment, poverty, and chronic illness. People with lower incomes and uninsured people are more likely to experience transportation barriers. Limited public transportation infrastructure in a neighborhood leaves people without access to vehicles lacking in reliable transportation options. Offering free or reimbursed public transit or taxi costs, connecting individuals to transportation, and policies that improve the safety and accessibility of sidewalks and bike lanes can help improve transportation access.

Lack of social and emotional support among adults

:   Positive relationships and interactions with family, friends, co-workers, and community members can have a protective impact on individual health and well-being, and these relationships can also help mitigate the negative impacts of challenges that people face (e.g., living in an unsafe neighborhood, trouble affording housing or food). Past research shows that people with high levels of social isolation (lack of interactions with others or the wider community) are at higher risk of developing chronic conditions (e.g., depression, cardiovascular disease, hypertension) and have a higher risk of mortality.

count_ntm_stops

:   Count of transit stops reported to National Transit Map as of 1/2023

stops_per_capita

:   Transit stops per 1000 people

stops_per_sqmile

:   Transit stops per square mile

census_tract_area20

:   Census land area, square miles

RUCA1

:   Primary RUCA Code

    | **Code** | **Classification description** |
    |----------------|--------------------------------------------------------|
    | 1 | Metropolitan area core: primary flow within an urbanized area (UA) |
    | 2 | Metropolitan area high commuting: primary flow 30% or more to a UA |
    | 3 | Metropolitan area low commuting: primary flow 10% to 30% to a UA |
    | 4 | Micropolitan area core: primary flow within an urban cluster of 10,000 to 49,999 (large UC) |
    | 5 | Micropolitan high commuting: primary flow 30% or more to a large UC |
    | 6 | Micropolitan low commuting: primary flow 10% to 30% to a large UC |
    | 7 | Small town core: primary flow within an urban cluster of 2,500 to 9,999 (small UC) |
    | 8 | Small town high commuting: primary flow 30% or more to a small UC |
    | 9 | Small town low commuting: primary flow 10% to 30% to a small UC |
    | 10 | Rural areas: primary flow to a tract outside a UA or UC |
    | 99 | Not coded: Census tract has zero population and no rural-urban identifier information |

RUCA2

:   Secondary RUCA codes, 2010

    | **Code** | **Classification description** |
    |--------------|----------------------------------------------------------|
    | 1 Metropolitan area core: primary flow within an urbanized area (UA) |  |
    | 1.0 | No additional code |
    | 1.1 | Secondary flow 30% to 50% to a larger UA |
    | 2 Metropolitan area high commuting: primary flow 30% or more to a UA |  |
    | 2.0 | No additional code |
    | 2.1 | Secondary flow 30% to 50% to a larger UA |
    | 3 Metropolitan area low commuting: primary flow 10% to 30% to a UA |  |
    | 3.0 | No additional code |
    | 4 Micropolitan area core: primary flow within an urban cluster of 10,000 to 49,999 (large UC) |  |
    | 4.0 | No additional code |
    | 4.1 | Secondary flow 30% to 50% to a UA |
    | 5 Micropolitan high commuting: primary flow 30% or more to a large UC |  |
    | 5.0 | No additional code |
    | 5.1 | Secondary flow 30% to 50% to a UA |
    | 6 Micropolitan low commuting: primary flow 10% to 30% to a large UC |  |
    | 6.0 | No additional code |
    | 7 Small town core: primary flow within an urban cluster of 2,500 to 9,999 (small UC) |  |
    | 7.0 | No additional code |
    | 7.1 | Secondary flow 30% to 50% to a UA |
    | 7.2 | Secondary flow 30% to 50% to a large UC |
    | 8 Small town high commuting: primary flow 30% or more to a small UC |  |
    | 8.0 | No additional code |
    | 8.1 | Secondary flow 30% to 50% to a UA |
    | 8.2 | Secondary flow 30% to 50% to a large UC |
    | 9 Small town low commuting: primary flow 10% to 30% to a small UC |  |
    | 9.0 | No additional code |
    | 10 Rural areas: primary flow to a tract outside a UA or UC |  |
    | 10.0 | No additional code |
    | 10.1 | Secondary flow 30% to 50% to a UA |
    | 10.2 | Secondary flow 30% to 50% to a large UC |
    | 10.3 | Secondary flow 30% to 50% to a small UC |
    | 99 Not coded: Census tract has zero population and no rural-urban identifier information |  |
